Dual host-parasite transcriptomes of apicomplexan Eimeria falciformis and its natural mouse host

### 2016-09-05 manuscript/bmc_article.tex
manuscript: changes in results (adjustments) and addition of heatmap introduction. GO-discussion not included

%%% Load packages
\usepackage{amsthm,amsmath}
\RequirePackage{natbib}
%\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
\RequirePackage{hyperref}
\usepackage[utf8]{inputenc} %unicode support
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
%\usepackage{siunitx}
\usepackage[colorlinks]{hyperref} % To color blocks of text 
%\usepackage[dvipsnames]{xcolor} % More color options (see Wikibooks)  
\usepackage{changepage}
\usepackage{booktabs} % For prettier tables
\usepackage{float} % To control positions of e.g. tables or figures
%\usepackage[section]{placeins}
\usepackage{graphicx}
\usepackage[flushleft]{threeparttable}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% 	OVERVIEW READS PER SAMPLE, LAYOT AS EXPERIMENTAL OVERVIEW
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

We performed mRNA sequencing of caecum tissue from mice infected with
the apicomplexan parasite \textit{E. falciformis}. Oocysts and
sporozoites were included as ``environmental'' stages and processed
``\textit{in vitro}''. To follow the lifecycle of the parsite, we
compare different timepoints after infection.

We additonally used different mouse strains which displayed different
immunocompetence in infection trials to assess the influence host
immunocompetence on parasite development (figure 1a,b). In these
experiments we analysed RAG-1-/- mice and compared them to the
parental C57BL6 mouse line. RAG knockout mich lack adaptive immune
response du to the absence of mature B and T lymphocytes.  All three
mouse strains were used for infections of both naive and previously
challenge infected animals.

Concerning basic phenotyping, differences in oocyst output were
visible both between the immunocompetent (NMRI and C57BL6) and
immunodeficient (RAG-1-/-) host strain, but also between the two
similarly immunocompetent (NMRI and C57BL6) strains (figure
1a). Surprisingly this finding was consistent in both in primary and
challenge infection. Similarly development of intestinal stages as
assayed by quantitative reverse transcription PCR (RT qPCR) also
... ADD HERE (figure 1b).

We thus aim to analyse the lifecycle of \textit{E. falciformis} under
the influence of (most likely the innate or cell autonomouse) immune
respons at on early stages of infection. We used an experimental
design comparing infections at 5dpi for all possible experimental
conditions (all three mouse strains in naive and challenge infection),
and analyse additional infections at 3dpi (figure 1c).

In our RNAseq experiment we worked with two biological replicates for
all but two conditions (with one and three replicates,
respecitively). For each replicate sample we enriched caecum tissue
from three infected mice for epithelial cells, extracted mRNA and
constructed sequencing libraries. These libraries were sequenced on
several lanes of Illumina GAIIX and HiSeq machines and mapped to both
mouse and parasite genome simulatneously to avoid spurious assignments
of read in ultra conserved genomic regions.

The number of total read mappings obtained for individual replicates
ranged from 11,479,260 (sample NMRI\_sporozoites\_rep2) to 145,033,326
(NMRI\_2ndInf\_5dpi\_rep1). As samples and idividual replicates were
sequenced in batches to different depth and using different
instrumentation (Table 1) we performed quality controls confirming the
abscence of batch effects influencing analysis and quality of our
results (additional file xyc). The results of these multivariate
analyses in relation to biological interpretation are described
further below.

\section*{Parasite and host transcriptomes can be assessed in parallel}

A maximum of 86\% (sample NMRI\_1stInf\_7dpi\_rep1) and a minimum of
0.0326\% (sample NMRI\_2ndInf\_3dpi\_rep2) of mapped reads could be
assigened to the \textit{E. falciformis} genome in samples considered
infected (table 1).

We excluded samples NMRI\_2nd\_3dpi\_rep1 and NMRI\_2nd\_5dpi\_rep2
due to an exptremely low parasite contribution to the overall
transcriptome. Technically this exclusion made it possible to obtain
read counts in agreement with a negative bionmial distributions (see
adddtional file x). Judging from their mRNA output those samples can
be considered non-infected, as both samples occured in a challenge
infection we assume that the infection was cleard or reduced to a
non-detectable output in those animmals.

One sample (NMRI\_1stInf\_0dpi\_rep1) was additonally excluded because
the uninfected control showed an unexpected mapping of reads to the
\textit{E. falciformis} genome. We consider this together with the
unsuccessfully infected samples above to display an undecided state of
infection.

On the other end of the spectrum mice not previously immunized
featured an mRNA ``holo-transcriptome'' at 7dpi with the majority of
mRNA originating from the parasite as opposed to the host (Figure
1). The overall mRNA output of the sampled caecum tissue at this point
of infection is dominated by the parasite with proportional mRNA
abundance of 77\% (in NMRI\_1stInf\_7dpi\_rep2) and even 92\% (in
NMRI\_1stInf\_7dpi\_rep1).

The total number of sequencing reads obtained as well as those reads
mapped to the transcriptome of either \textit{E. falciformis} and the
mouse are indicated in Table 1 for all replicates. We additonally
assessed the number of genes considered ``expressed'', further
analysis and discussion considers the possibility that especially for
\textit{E. falciformis} the number of reads might have been
insufficient to fully normalize the dataset of the samples with the
lowest representation against those with the hightest.

\section*{The mouse transcriptome changes in the course of infection infection}

Statistical testing for differetial expression between infected and
uninfected mice, revealed changes in mRNA abundance becoming more
pronounced (involving more genes) at later post infection (especially
at 7dpi; Table 2).

The observed changes between controls and 7dpi were in agreement with
previously published microarry data. Fold-Change data obtained from
mice at 6dpi on Agilent microarrays (Schmidt et al 2012) or 7dpi (our
RNAseq data) against uninfected controls shows a strong correlateion
(Spearman's \rho = 0.74; figure 2a). Considering both and technical
differences between the tow methods and the biological changes along
the lifecycle this confirms the adequacy of our methods for assessing
the host transcriptome.

325 mouse genes were considered differently expressed (DE; FDR<1\%)
between uinfected and 3dpi, 1,804 between controls and 5dpi and 2,711
genes between 0dpi and 7dpi. This lead to combined set of 3453 genes
reacting to infection (figure 2bi).  DE genes early in infection (3dpi
and and 5dpi) were not simply a subset of genes becoming DE later in
infection (7dpi). To further validate this pattern of a distinct
response in early infection we performed hierarchical clustering on
(the union of) mouse genes DE between different dpi (figure 2c).

The ponounced changes later in infection are reflecting the expected
onset of an adaptive immune response. This is underlined by GO-terms
such as ``antigen binding'' and ``immunoglobulin receptor binding''
(molecular function) ``immune system process'' and ``adaptive immune
response'' (biological process) highly enriched in Cluster 5 of figure
2c.

Particular ``regulation of cell death'' and ``regulation of apoptotic
process'' (BP)

``UDP-N-acetylglucosamine-lysosomal-enzyme'' (MF)

@@

General changes upon infection are a downregulation of genes involved
``lipid modification'' or more detailed ``phosphatidylinositol
phosphorylation''. These changes are similarly pronounce in the
immunocompetent strains and RAG-/- immunodefficient animals (cluster 3
of fibure 2c) (see supplementary table X).  The presence of these
changes immunocompromized animals invite the speculation that those
genes are independent of a potential later adaptive immune response.

Changes not detectable in RAG-/- immunodefficient animals (cluster 3
of fibure 2c) but otherwise observed in general infection are
characterized by and enrichment to genes involved in ``positive
regulation of calcineurin-NFAT'', ``blood coagulation'' and ``inositol
phosphate-mediated signaling'' (see supplementary table X). The
abscence of these changes in RAG-/- mice allow us to speculate that
those genes might be involved in a very early onlset of adaptive
immune response.

@@

We dedtected further differences between infections of naive and
challenge infected mous hosts , especially at 7 dpi. These genes are
characterized by...
%% We should't limit ourselfs to testing the clusters only! But
%% discuss the cluster first if you got difficulties coding the
%% enrichment tests.

\section*{The lifecycle of Eimeria is characterized by major changes in mRNA abundance}

Table 2 lists pairwise sample comparisons and number of genes with
differently abundant mRNAs per comparison.  For
\textit{E. falciformis} no mRNAs are significantly differently
abundant between first and second infection.

Changes throughout the intr-host lifecycle can be clearly categorized
in an early (3 and 5 dpi) and late (7dpi) phase. Sporozoites show
expression profiles similar more similar to early infection stages
while oocysts show distinctly different mRNA abundance profile (MDS
plot .

We applied hierarchical clustering on \textit{E. falciformis} mRNAs
which were differently abundant at different dpi, including
comparisons with oocysts and sporozoites.

Samples from 7dpi form a separate cluster (NMRI mice only), not
distinguishing between first and challenge infection.

@@

\subsubsection*{Data overview}
In total, xx reads were sequenced. For each sample, mouse read numbers were $10^7$ - $10^8$
Quality filters applied resulted in xx reads being removed from further analysis

>>>>

%%% Load packages
\usepackage{amsthm,amsmath}
%\usepackage{asmath}
\RequirePackage{natbib}
%\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
\RequirePackage{hyperref}
\usepackage[utf8]{inputenc} %unicode support
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
%\usepackage{siunitx}
\usepackage[colorlinks]{hyperref} % To color blocks of text 
%\usepackage[dvipsnames]{xcolor} % More color options (see Wikibooks)  
\usepackage{changepage}
\usepackage{booktabs} % For prettier tables
\usepackage{float} % To control positions of e.g. tables or figures
%\usepackage[section]{placeins}
\usepackage{graphicx}
\usepackage[flushleft]{threeparttable}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% 	OVERVIEW READS PER SAMPLE, LAYOT AS EXPERIMENTAL OVERVIEW
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% 	OVERVIEW READS PER SAMPLE, LAYOT AS EXPERIMENTAL OVERVIEW
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

We performed mRNA sequencing of caecum tissue from mice infected with
the apicomplexan parasite \textit{E. falciformis}. Oocysts and
sporozoites were included as ``environmental'' stages and processed
``\textit{in vitro}''. To follow the life cycle of the parasite, we
compare different time-points after infection.

We additionally used different mouse strains which display different
immunocompetence in infection trials measured by oocyst output to assess the influence of host
immunocompetence on parasite development (figure 1a,b). In these
experiments we analysed mice lacking the Recombination activating gene 1, Rag1, and compared them to the
parental C57BL/6 mouse line. Rag1 knock-out mice lack adaptive immune
response due to the absence of mature B and T lymphocytes.  All three
mouse strains were used for infections of both naive mice and previously
infected animals (onward referred to as "challenged").

Basic phenotyping showed differences in oocyst numbers
between the immunocompetent (C57BL/6) and
immunodeficient (Rag1-/-) host strain in infection of naive mice and 
in challenge infections. Immunocompetent NMRI mice were infected with a higher dosis  
of sporulated oocysts, and a drastic reduction of oocysts in faeces was seen in challenged mice
compared to naive mice (Figure 1a). Oocyst numbers in faeces peaked on days 8-9 and 
all mice had cleared the infection by day 14. Development of \textit{E. falciformis} intestinal stages is reflected by 
the percentage of parasite reads sequenced per time-point post infection (Figure 1b) and this was confirmed by quantitative reverse transcription PCR (RT qPCR) of parasitic 18S  (Figure 1c).

We thus use dual RNA-seq to analyse the life cycle of \textit{E. falciformis} under
the influence of host immune responses at early stages of infection. We used an experimental
design which allows to compare infections at 5 days post infection (dpi) for all experimental
conditions (NMRI, C57BL/6 and Rag1-/- mouse strains in naive and challenge infection). In addition,
additional time-points 3 dpi and 7 dpi were analysed from NMRI mice (Figure 1d).

\section*{Parasite and host dual transcriptomes can be assessed in parallel}
In this RNA-seq experiment, each replicate sample was enriched for caecum epithelial tissue which was pooled 
from three mice. mRNA was extracted and
sequencing libraries prepared. Two biological replicates were used for
all but two conditions (with one and three replicates, respectively). These libraries were sequenced on
several lanes of Illumina GAIIX (13 samples) and HiSeq machines (14 samples) and mapped to both
mouse and parasite genomes simultaneously to avoid spurious assignments
of reads in ultra conserved genomic regions. 
As samples and individual replicates were sequenced in batches to different depth and using different
instrumentation (Table 1) we performed quality controls (additional files xyz). These confirm
the absence of batch effects influencing analysis and quality of 
results.

Total numbers of sequenced reads as well as reads
mapped to either the \textit{E. falciformis} genome or the
mouse genome are indicated in Table 1 for all replicates. The number of total read mappings 
for individual replicates ranged from 11,479,260 (sample NMRI\_sporozoites\_rep2) to 145,033,326
(NMRI\_2ndInf\_5dpi\_rep1). At the latest time-point samples, 7 dpi, the overall mRNA
output of the sampled caecum tissue is dominated by parasite material with proportional mRNA
abundance of 77\% (in NMRI\_1stInf\_7dpi\_rep2) and 92\% (in
NMRI\_1stInf\_7dpi\_rep1). (Figure 1b).

\section*{Exclusion of samples with uncertain infection status}
A maximum of 86\% (sample NMRI\_1stInf\_7dpi\_rep1) and a minimum of
0.0326\% (sample NMRI\_2ndInf\_3dpi\_rep2) of mapped reads could be
assigned to the \textit{E. falciformis} genome in samples considered
infected (Table 1). We excluded samples NMRI\_2nd\_3dpi\_rep1 and NMRI\_2nd\_5dpi\_rep2
due to low parasite contribution to the overall transcriptome. 
Technically, this exclusion made it possible to obtain read counts in agreement with a 
negative binomial distribution (see additional file x). It is also likely that the number of reads in the excluded samples would have been
insufficient to fully normalise these datasets to those with the highest parasite contributions.
From a biological point of view, both excluded are samples from challenge
infection and we assume that the infection was cleared or reduced to a
non-detectable level. One sample (NMRI\_1stInf\_0dpi\_rep1) was excluded because
the uninfected control showed unexpected mapping of reads to the
\textit{E. falciformis} genome. We consider the three excluded samples to display an uncertain 
state of infection.

\section*{The mouse transcriptome changes in the course of infection}
Statistical testing for differential expression between infected and
uninfected mice revealed changes in mRNA abundance becoming more
pronounced (involving more genes) at later time-points post infection (especially
at 7 dpi; Table 2).

The observed changes between controls and 7 dpi were in agreement with
previously published microarry data. Fold-Change data obtained from
\textit{E. falciformis} infected mice at 6 dpi on Agilent microarrays 
(Schmidt et al 2012) or 7 dpi (our RNAseq data) against uninfected controls
show a strong correlation (Spearman's \rho = 0.74; Figure 2a). Considering both biological 
(e.g. exact time-points) and technical differences between the two methods this comparison 
confirms the adequacy of using dual RNA-seq for assessing the host transcriptome.

For mouse, 325 mRNAs were considered differently abundant (DA; FDR<1\%)
between controls and 3 dpi, 1,804 mRNAs between controls and 5 dpi and 2,711
mRNAs between controls and 7 dpi. This lead to a combined set of 3,453 genes
responding to infection (Figure 2bi). DA mRNAs early in infection (3 dpi
and 5 dpi) were not simply a subset of genes DA later in
infection (7 dpi), but instead, the transcriptional profile of the mouse 
changes throughout the time-points tested here. To further validate this pattern of a distinct
response in early infection we performed hierarchical clustering on
the (union of) mouse genes DA between different time-points post infection (Figure 2c).

By hierarchical clustering of mouse samples, three main sample clusters formed, “main clusters” 
(dendrogram of columns at top of Figure 2c). Rag1-/- mice cluster with control samples, and there is no clear distinction between infected and non-infected Rag1-/- samples, confirming the immune deficiency phenotype in Rag1-/-. Control samples are defined by high abundance of transcripts in gene cluster 3 (yellow, clusters of rows on the left) and partly high abundance in cluster 4 (red). Immunodeficient Rag1-/- samples share this profile with control samples in gene cluster 4, and this cluster has low abundance in (most) other infection samples. This suggests that these are genes which become down-regulated in immune competent mice upon infection. 
%%Two naive 5 dpi samples to the left cluster separately and diverge from this and most other patterns. High in infection, but does not fit with third NMRI replicate....
Three samples of challenged mice from 3, 5, and 7 dpi display high abundance of transcripts in gene cluster 6 (blue). The same genes have a low abundance in two naive, immune competent mouse samples from 5 dpi. Genes in cluster 6 are therefore particularly interesting for investigating the role of adaptive immunity in this infection (although sample NMRI\_2ndInf\_5dpi\_rep1 has a different profile). 
%% GO terms in cluster 6 discuss

Transcripts in gene clusters 3 and 4 are lowly abundant among all samples in the middle main cluster. Among those samples, the NMRI mouse samples display a trend towards high abundance in clusters 1, 2, 5 (green) and 7. Especially in the two late infection samples (7 dpi) a high transcript abundance is seen in clusters 1 and 5. The leftmost main sample cluster containing immunocompetent 5 dpi samples is characterized by low transcript abundance in cluster 6, and high abundance in cluster 7.

The pronounced changes later in infection reflect the expected
onset of an adaptive immune response. This is underlined by enriched Gene Ontology, GO, terms
in gene clusters with pronounced profiles late in infection (Figure 2c). Terms such as "antigen binding" and "immunoglobulin receptor binding" (molecular function, MF), and "immune system process" and "adaptive immune
response" (biological process, BP) are highly enriched in Cluster 5 and confirm an activated adaptive immune response
at this time-point.

Particular "regulation of cell death" and "regulation of apoptotic process" (BP)

"UDP-N-acetylglucosamine-lysosomal-enzyme" (MF)

@@

General changes upon infection are a down-regulation of genes involved
"lipid modification" or more detailed "phosphatidylinositol
phosphorylation". These changes are similarly pronounce in the
immunocompetent strains and Rag1-/- immunodefficient animals (cluster 3
of fibure 2c) (see supplementary table X).  The presence of these
changes immunocompromised animals invite to the speculation that those
genes are independent of a potential later adaptive immune response.

Changes not detectable in Rag1-/- immunodefficient animals (cluster 3
of fibure 2c) but otherwise observed in general infection are
characterised by and enrichment to genes involved in "positive
regulation of calcineurin-NFAT", "blood coagulation" and "inositol
phosphate-mediated signaling" (see supplementary table X). The
absence of these changes in Rag1-/- mice allow us to speculate that
those genes might be involved in a early onset of adaptive
immune responses.

@@

We detected further differences between infections of naive and
challenge infected mouse hosts, especially at 7 dpi. These genes are
characterized by...
%% We should't limit ourselfs to testing the clusters only! But
%% discuss the cluster first if you got difficulties coding the
%% enrichment tests.

\section*{The life cycle of E. falciformis is characterized by major changes in mRNA abundance}

Table 2 lists pairwise sample comparisons and number of genes with
differently abundant mRNAs per comparison. We applied hierarchical clustering on 
\textit{E. falciformis} mRNAs which were differently abundant at different time-points 
post infection, including comparisons with oocysts and sporozoites. 
Upon hierarchical clustering (Figure 3c) changes throughout the parasite's intracellular life cycle 
can be categorised into an early (3 and 5 dpi) and late (7dpi) phase. 
Mouse immune status has no apparent influence on clustering of
parasite samples. Sporozoites show expression profiles similar to early infection stages
while oocysts show a distinct mRNA abundance profile (MDS plot...).

Parasite gene clusters shown on the left side (colour-coded and numbered 1 - 7) can be used 
to characterise clusters of samples. Late in infection, a profile of high mRNA abundance in clusters 2 and 7 (mint green and purple) is seen. Also parasite genes in cluster 6 (blue) appear characteristic  by displaying low mRNA abundance in late stages of infection. 
 One late challenge infection sample deviates from this pattern together with an early challenge infection sample. These two samples are, apart from controls and excluded samples, the ones with the least number of detected parasite reads (1836 and 1580, respectively), which taken together suggests that infections had been cleared before sampling in these challenge infected mice.

The early infection sample cluster (12 samples in center of heatmap) is characterised by a high abundance of transcripts in clusters 3 (yellow) and 6, as well as a below average abundance in clusters 1 (pink), 2, and 5 (green). 
Sporozoite samples have highly abundant transcripts in cluster 4 (red) and low abundance of 
transcripts in gene cluster 7. The latter is shared with oocyst samples. The oocyst samples, 
in addition, have low abundance of transcripts in gene clusters 2, 3, 4, 6 (and 7) and highly 
abundant transcripts in clusters 1 and 5.

Taken together, sample and gene clustering indicates that genes in clusters 2, 6 and 7 might be 
useful to characterise assumed gametocytes in 7 dpi samples. Genes in cluster 4 are candidates 
for sporozoite specific genes, and clusters 1 and 5  for oocyst specific genes in E. falciformis.

@@
